---
title: "Effect of FUS-KI on RNA stability"
author: "Katharina Hembach"
date: "11/12/2020"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), warnings = FALSE)
```

#### Load libraries
```{r load-libs, message = FALSE}
library(here)
library(dplyr)
library(ComplexHeatmap)
library(eulerr)
library(ggplot2)
library(stringr)
library(ggrepel)
library(rtracklayer)
library(viridis)
library(tibble)
library(biomaRt)
library(limma)
library(cowplot)
library(factoextra)
library(NbClust)
library(cluster)
```


## Load data

```{r file-names}
se <- here("output", "outputR", "edgeR_dge.rds")

## Manually filtered list of target genes (genes on the wrong strand are 
## removed)p; results from RNA-seq of 6months old FUS-KI and WT mice is 
## included in the table
fus_targets <- "/home/Shared_taupo/data/seq/sonu_CLIPseq/clip_March2018/analysis/deduplicated/MA_plot_selection/top_peaks_loess_adjM_pValue_1e-05_gene_anno_manually_filtered_cluster_location_classified_RNAseq_6m.txt"
fus_peaks_bed <- "/home/Shared_taupo/data/seq/sonu_CLIPseq/clip_March2018/analysis/deduplicated/MA_plot_selection/selected_peaks/selected_peaks_SNS_pValue_1e-05.bed"
## annotated fus peaks
fus_peaks_rds <- "/home/Shared_taupo/data/seq/sonu_CLIPseq/clip_March2018/Rmd/rds/loess_peaks_sns_pValue_1e-05.RDS"

gtf_file <- "/home/Shared_taupo/data/annotation/Mouse/Ensembl_GRCm38.90/gtf/Mus_musculus.GRCm38.90.gtf"
fig_dir <- here("analysis", "figures")
```


```{r load-data, dependson = "file-names"}
se <- readRDS(se)
## List of SummarizedExperiment objects (gene/transcript level)
se
## Get gene-level SummarizedExperiment object
sg <- se$sg
metadata <- colData(sg)
##log2 counts per million
logcpms <- assay(sg, "logcpm")

targets <- read.table(fus_targets, header = TRUE)
```

# EdgeR results for the four tested contrasts

We tested for differences in gene expression between the KI and the WT samples per time points. There are four timepoints: 0, 8, 12 and 24 hours.
The edgeR results are stored in the `rowData` slot of the `sg` object. We keep all protein coding genes.

```{r get-edgeR-res}
rd <- rowData(sg)
contrasts <- c("groupKI_0h-groupWT_0h", "groupKI_8h-groupWT_8h", 
               "groupKI_12h-groupWT_12h", "groupKI_24h-groupWT_24h")
res_all <- list("0h" = rd[,"edgeR:groupKI_0h-groupWT_0h"],
            "8h" = rd[,"edgeR:groupKI_8h-groupWT_8h"],
            "12h" = rd[,"edgeR:groupKI_12h-groupWT_12h"],
            "24h" = rd[,"edgeR:groupKI_24h-groupWT_24h"])
```

We only keep the protein coding and lincRNA genes for our analysis. 

```{r}
gtf <- import(gtf_file)
genes <- gtf[gtf$type == "gene"]
protein_genes <- gtf[gtf$gene_biotype == "protein_coding"]

## we annotate the gene biotype
res_all <- lapply(res_all, function(x) {
  m <- match(x$gene_id, genes$gene_id)
  x$gene_biotype <- NA
  x$gene_biotype[!is.na(m)] <- genes[m[!is.na(m)]]$gene_biotype
  x
})

res_all <- lapply(res_all, function(x) {
  x[x$gene_biotype %in% c("protein_coding", "lincRNA"),]
})
```


We count the number of DE genes and overlap them with the lists of FUS targets from the CLIP-seq.

```{r overlap-FUS-targets}
res <- lapply(res_all, function(x) x %>% as.data.frame %>% 
                dplyr::filter(FDR <= 0.05))
lapply(res, nrow)
## overlap the DE genes with the FUS targets
for (i in names(res)) {
  dat <- res[[i]]
  dat$gene_name %in% targets$gene_name %>% table %>% print
  print(nrow(dat))
}
```


# PCA plot

We use the logCPM values for PCA, because they are normalized for library size and transcript length.

```{r PCA-plot}
## Filter out all genes that were not used for the DE analysis (they are NA) and that we removed above
keep <- !is.na(res_all[[1]]$PValue) 
gene_ids <- res_all[[1]]$gene_id[keep]
g <- rownames(rowData(sg)[rowData(sg)$gene_id %in% gene_ids,]) 
logcpms_part <- logcpms[g,]


## Since PCA can be slightly problematic with high dimensional data, we first 
## use select only the 1000 genes showing the highest variance.
ntop = 1000
pvars <- rowVars(logcpms_part)
select <- order(pvars, decreasing = TRUE)[seq_len(min(ntop, length(pvars)))]

## We need genes as columns --> transpose
pca <- prcomp(t(logcpms_part[select,]))

#Determine the proportion of variance of each component
#Proportion of variance equals (PC stdev^2) / (sum all PCs stdev^2)
proportionvariances <- ((pca$sdev^2) / (sum(pca$sdev^2)))*100
proportionvariances <- round(proportionvariances, digits = 0)

## PCA plot
pca_x <- as.data.frame(pca$x) %>% tibble::rownames_to_column(var = "names") %>%
    dplyr::full_join(data.frame(colData(sg)), by = "names")

ggplotPCA <- function(dat, x = "PC1", y = "PC2", color = "group", 
                      shape = "genotype", 
                      color_pal = brewer.pal(4, name = "GnBu"),
                      var_explained_x = 0, var_explained_y = 0){
  ggplot(dat, aes_string(x = x, y = y, color = color, shape = shape)) +
    geom_point(size = 4, alpha = 0.8) +
    theme_bw(base_size = 18) +
    scale_color_manual(values = color_pal) + 
    theme(aspect.ratio = 1) + 
    labs(x = paste0(x, "(", var_explained_x, "%)"), 
         y = paste0(y, "(", var_explained_y, "%)"))
}

## custom colors for Magda
pca_x$genotype <- factor(pca_x$genotype, levels = c("WT", "KI"))
pca_x$time <- factor(pca_x$time, levels = c("0h", "8h", "12h", "24h"))

p <- ggplotPCA(pca_x, x = "PC1", y = "PC2", shape = "genotype", 
          color = "time", 
          color_pal = inferno(4, begin = 0.2, end = 0.8, direction = -1), 
          var_explained_x = proportionvariances[1], 
          var_explained_y = proportionvariances[2])
p

p <- ggplotPCA(pca_x, x = "PC3", y = "PC4", shape = "genotype", 
          color = "time", 
          color_pal = inferno(4, begin = 0.2, end = 0.8, direction = -1), 
          var_explained_x = proportionvariances[3], 
          var_explained_y = proportionvariances[4])
p
```



# Volcano plot

volcano plot for each of the timepoints.

```{r volcano-prep}
## select the ones we are interested in:
comp <- names(rowData(sg))[9:12]

## Determine the maximal logFC range in all comparisons
x_range <- 0
y_range <- 0
for (i in names(res_all)){
  dat <- res_all[[i]]
  dat <- dat[!is.na(dat$PValue),]
  x_range<- max(x_range, max(abs(dat$logFC)))
  y_range <- max(y_range, max(-log10(dat$FDR)))
}
x_range
y_range
```


```{r "volcano-plot-function"}
## Volcano plot with labeled genes (top p-value or logFC)
ggplotVolcano <- function(dat, title, outfile, alpha = 0.8, FUS_targets = NULL, 
                          color_targets = FALSE, gene_labels = NULL,
                          label_top = FALSE,
                          label_coord = 1.5, 
                          x_range = NULL, y_range = NULL){

  dat <- dat[!is.na(dat$PValue),]
  dat$mlog10FDR <- -log10(dat$FDR)
  df <- data.frame(logFC = dat$logFC, mlog10FDR = dat$mlog10FDR, 
                      FDR = dat$FDR, gene = dat$gene_name)
  
  p <- ggplot(df, aes(x = logFC, y = mlog10FDR))
   
  if(is.null(FUS_targets)){
    cols <- as.factor(ifelse(df$FDR <= 0.05, 
                             ifelse(df$logFC > 0, "FDR \u2264 0.05 & logFC > 0",  
                                    "FDR \u2264 0.05 & logFC < 0"), "FDR > 0.05"))
    # repel <- df %>% filter(FDR <= 0.05) %>% arrange(desc(mlog10FDR))
    p <- p + geom_point(alpha = alpha, aes(color = cols)) +
      scale_color_manual(values = c("FDR \u2264 0.05 & logFC > 0" = "#882255", 
                                    "FDR \u2264 0.05 & logFC < 0" = "#117733",
                                    "FDR > 0.05" = "darkgrey"))
  } else {
    # repel <- df %>% filter(gene %in% FUS_targets) %>% 
    #   arrange(desc(mlog10FDR))
    if (color_targets){
     df$cols <- as.factor(ifelse(df$gene %in% FUS_targets, "FUS target",
                              ifelse(df$FDR <= 0.05, 
                                     ifelse(df$logFC > 0, "FDR \u2264 0.05 & logFC > 0",  
                                            "FDR \u2264 0.05 & logFC < 0"), "FDR > 0.05")))
      p <- p + geom_point(data = df %>% dplyr::arrange(cols), 
                          aes(color = cols), alpha = alpha) + 
        scale_color_manual(values = c("FUS target" = "steelblue3", 
                                      "FDR > 0.05" = "darkgrey", 
                                      "FDR \u2264 0.05 & logFC > 0" = "#882255", 
                                      "FDR \u2264 0.05 & logFC < 0" = "#117733"))
     } else{
      p <- p + geom_point(data = df %>% filter(!gene %in% FUS_targets), 
                   aes(color = "darkgrey"), alpha = alpha) + 
        geom_point(data = df %>% filter(gene %in% FUS_targets),
               aes(color = "red"), alpha = alpha) +
      scale_color_manual(name = "FUS targets", labels = c("FALSE", "TRUE"), 
                         values = c("darkgrey", "red"))
    }
  }
  p <- p + ggtitle(title) +
    theme_bw() +
    theme(text = element_text(size = 16)) + 
    ylab("-log10(FDR)") 
  
  if(!is.null(gene_labels)){
    repel <- df %>% filter(gene %in% gene_labels) %>% 
      arrange(desc(mlog10FDR))
    p <- p + geom_text_repel(data = repel %>% filter(logFC > 0), 
                           aes(label = gene), xlim = c(label_coord, NA), 
                           segment.color = "grey30", show.legend = FALSE) +
      geom_text_repel(data = repel %>% filter(logFC < 0), 
                      aes(label = gene), xlim = c(NA, -label_coord), 
                      segment.color = "grey30", show.legend = FALSE)
  } else if (label_top){
    ## top FDR
    repel <- df %>% filter(FDR <= 0.05) %>% arrange(desc(mlog10FDR)) %>% 
      dplyr::slice(1:10)
    ## top logFC
    repel <- df %>% arrange(desc((logFC))) %>%
      dplyr::slice(1:10) %>% rbind(repel)
    repel <- df %>% arrange(logFC) %>%
      dplyr::slice(1:10) %>% rbind(repel)
    repel <- unique(repel)
    
    p <- p + geom_text_repel(data = repel %>% filter(logFC > 0), 
                             aes(label = gene), xlim = c(label_coord, NA), 
                             segment.color = "grey30", show.legend = FALSE) +
      geom_text_repel(data = repel %>% filter(logFC < 0), 
                      aes(label = gene), xlim = c(NA, -label_coord), 
                      segment.color = "grey30", show.legend = FALSE)
  }
    
  ## x and y-axis range
  if(!is.null(x_range) & !is.null(y_range)){
      p <- p + coord_cartesian(xlim = c(-x_range, x_range), ylim = c(0, y_range))
  }
  
  ## draw line at FDR cutoff and end x = 0 
  p <- p + theme(aspect.ratio = 1) + 
    geom_hline(yintercept = -log10(0.05), linetype="dashed", color = "black")  +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black")
  
  ## remove legend title
  p <- p + theme(legend.title=element_blank()) + xlab("log2 fold change")
   
  print(p)
  ggsave(filename = outfile, plot = p, width = 8, height = 6, type = "cairo", 
         dpi = 500)
}
```

```{r volcano-plots1,  warning = TRUE}
for (i in names(res_all)){  
  file_name <- file.path(fig_dir, paste0("volcano_", i, ".png"))
  ggplotVolcano(res_all[[i]], title = i, file_name, label_coord = 2.5, 
                x_range = x_range, y_range = y_range)
}
```

Label the top 15 genes with strongest logFC.
```{r volcano-top-label}
for (i in names(res_all)){  
  file_name <- file.path(fig_dir, paste0("volcano_", i, "_top_label.png"))
  ggplotVolcano(res_all[[i]], title = i, file_name, label_coord = 2.5, 
                x_range = x_range, y_range = y_range, label_top = TRUE)
}
```


## Volcano plot with labeled FUS target genes

```{r volcano-FUS-targets}
for (i in names(res_all)){  ## total
  file_name <- file.path(fig_dir, paste0("volcano_", i, "_FUS_targets.png"))
  ggplotVolcano(res_all[[i]], title = i, file_name, 
                FUS_targets = targets$gene_name, color_targets = TRUE, 
                alpha = 0.8,  label_coord = 2.5,  x_range = x_range, 
                y_range = y_range)
}
```


## Genes with strongest logFC

What are the DE genes with the highest logFC per timepoint?

```{r strongest-logFC}
lapply(res, function(x){
  x %>% as.data.frame %>% dplyr::arrange(desc(abs(logFC))) %>% head(n = 10)
})
```


# ORA 

Over representation analysis to characterize the sets of differentialy expressed genes.

The background are all genes with cpm > 1 in all WT 1 replicates.

```{r import-SNS-bg}
## all genes with >= 1 CPM in all replicates per group (logcpm 0 = cpm 1)
rna_wt <- logcpms %>% as.data.frame() %>% 
    dplyr::select(metadata[metadata$group == "WT_0h", "names"]) %>%
    rownames_to_column("gene_id") %>%
    dplyr::filter(., rowSums(dplyr::select(., -gene_id) >= 0) == (ncol(.)-1)) %>%
    tidyr::separate(gene_id, into = c("gene_id", "gene_name"), sep = "__")
dim(rna_wt)

rna_ki <- logcpms %>% as.data.frame() %>% 
    dplyr::select(metadata[metadata$group == "KI_0h", "names"]) %>%
    rownames_to_column("gene_id") %>%
    dplyr::filter(., rowSums(dplyr::select(., -gene_id) >= 0) == (ncol(.)-1)) %>%
    tidyr::separate(gene_id, into = c("gene_id", "gene_name"), sep = "__")
dim(rna_ki)

rna_bg <- c(rna_wt$gene_id, rna_ki$gene_id) %>% unique
length(rna_bg)
```


```{r convert-ensembl-ID-entrez}
## convert to Entrez IDs
ensembl <- useMart("ensembl")
ensembl <- useDataset("mmusculus_gene_ensembl", mart = ensembl)

if (file.exists(here("Rmd", "rds", "wt_bg_entrez.rds"))) {
  bg_wt <- readRDS(here("Rmd", "rds", "wt_bg_entrez.rds"))  
} else{
  ## convert to Entrez IDs
  bg_wt <- getBM(attributes = c("ensembl_gene_id","entrezgene_id"), 
                         values = rna_bg, mart = ensembl )
  bg_wt <- bg_wt[bg_wt$ensembl_gene_id %in% rna_bg, ]
  saveRDS(bg_wt, here("Rmd", "rds", "wt_bg_entrez.rds")) 
}
nrow(bg_wt)
```


```{r ORA}
ora_res <- list()

for (s in names(res)[2:4]) {
  up <- res[[s]] %>% dplyr::filter(FDR <= 0.05 & logFC > 0) %>% pull(gene_id)
  down <- res[[s]] %>% dplyr::filter(FDR <= 0.05 & logFC < 0) %>% pull(gene_id)
  
  direction <- c("up", "down")
  gene_lists <- list(up = up, down = down)
  for (i in direction) {
    name <- paste0(s, "_", i)
    ids <- gene_lists[[i]]
    
    if (file.exists(here("Rmd", "rds", paste0(name, "_entrez.rds")))) {
      entrez_ids <- readRDS(here("Rmd", "rds", paste0(name, "_entrez.rds")))  
    } else{
      ## convert the Ensembl Gene IDs to entrez Gene IDs with Biomart
      entrez_ids <- getBM(attributes = c("ensembl_gene_id","entrezgene_id"), 
                   values = ids, mart = ensembl )
      ## we only keep the genes from our initial list
      entrez_ids <- entrez_ids[entrez_ids$ensembl_gene_id %in% ids, ]
      saveRDS(entrez_ids, here("Rmd", "rds", paste0(name, "_entrez.rds"))) 
    }
    
    ## Match gene length to bg list
    m <- match(bg_wt$ensembl_gene_id, genes$gene_id)
    gene_length <- width(genes)[m]
    
    ## GO analysis with goana using gene length as covariate
    goana_res <- goana(de = unique(entrez_ids$entrezgene), species = "Mm", 
                       universe = bg_wt$entrezgene, covariate = gene_length)
    goana_res <- goana_res[order(goana_res$P.DE, decreasing = FALSE),]
    ora_res[[name]] <- goana_res
    
    ## write results
    write.table(topGO(goana_res, ontology = "CC", number = 100),
                here("analysis", "goana", paste0(name, "_goana_CC.txt")),
                sep = "\t", quote = FALSE)
    write.table(topGO(goana_res, ontology = "BP", number = 100),
                here("analysis", "goana", paste0(name, "_goana_BP.txt")),
                sep = "\t", quote = FALSE)
    write.table(topGO(goana_res, ontology = "MF", number = 100),
                here("analysis", "goana", paste0(name, "_goana_MF.txt")),
                sep = "\t", quote = FALSE)
  }
}
```

```{r print-ora-res}
for (s in names(ora_res)) {
  print(s)
  ## What are the most enriched annotation terms?
  print(topGO(ora_res[[s]], ontology = "CC", number = 10))
  print(topGO(ora_res[[s]], ontology = "BP", number = 10))
  print(topGO(ora_res[[s]], ontology = "MF", number = 10))
}
```



# Comparison of the DE genes and the FUS targets

## Upset plot
```{r upset-plot}
gene_list <- list("FUS targets" = targets$gene_name,
                  "8h" = unique(res[["8h"]]$gene_name),
                  "12h" = unique(res[["12h"]]$gene_name),
                  "24h" = unique(res[["24h"]]$gene_name))
comb_mat <- make_comb_mat(gene_list, mode = "distinct")

UpSet(comb_mat, set_order = c("FUS targets", "8h", "12h", "24h"), 
      c("steelblue", "#117733", "#882255")[comb_degree(comb_mat)])

## add intersection size to plot
# ht = draw(UpSet(comb_mat))
# od = column_order(ht)
# cs = comb_size(comb_mat)
# decorate_annotation("intersection_size", {
#     grid.text(cs[od], x = seq_along(cs), y = unit(cs[od], "native") + unit(2, "pt"), 
#         default.units = "native", just = "bottom", gp = gpar(fontsize = 8))
# })
```


## Venn diagrams
```{r venn-diagram}
## check if al gnees are unique
lengths(gene_list)
lapply(gene_list, function(x) length(unique(x)))


euler_diag <- euler(gene_list, shape = "circle")

eulerr_options(labels = list(fontsize = 20), 
               quantities = list(fontsize = 20, font = 2), 
               fills = list(alpha = 0.5),
               padding = unit(0.6, "lines"))

p <- plot(euler_diag, font=1, 
     quantities = TRUE, edges = list(lty = 2, col = "darkgrey"),
     alpha=0.8, lwd = 3, adjust_labels = TRUE)
print(p)
```

## List of overlapping genes
What are the FUS targets that are DE in the stability experiment?
```{r olap-genes}
res[["8h"]] %>% dplyr::filter(gene_name %in% targets$gene_name)
res[["12h"]] %>% dplyr::filter(gene_name %in% targets$gene_name)
res[["24h"]] %>% dplyr::filter(gene_name %in% targets$gene_name)
```

How many of them were DE in the 6 months SNS FUS-KI vs. WT RNA-seq?

```{r res-tables}
res_tables <- list("8h" = 
  res[["8h"]] %>% dplyr::filter(gene_name %in% targets$gene_name) %>% 
    left_join(targets, by = c("gene_id", "gene_name"), suffix = c(".stab", ".SNS")),
  "12h" = res[["12h"]] %>% dplyr::filter(gene_name %in% targets$gene_name) %>% 
    left_join(targets, by = c("gene_id", "gene_name"), suffix = c(".stab", ".SNS")),
  "24h" = res[["24h"]] %>% dplyr::filter(gene_name %in% targets$gene_name) %>% 
    left_join(targets, by = c("gene_id", "gene_name"), suffix = c(".stab", ".SNS")))

## number of DE genes in SNS 6months RNA-seq
lapply(res_tables, function(x) x$RNAseq_DE %>% table)

## write the results tables to file
for (n in names(res_tables)) {
  write.table(res_tables[[n]], file = here("analysis", paste0(n, "_DE_stability_FUS_targets.txt")),
              quote = FALSE, row.names = FALSE, sep = "\t")
}

## tables with all the DE genes from the stability experiment (most are not FUS targets)
res_all_tables <- list("8h" = res[["8h"]] %>% 
left_join(targets, by = c("gene_id", "gene_name"), suffix = c(".stab", ".SNS")),
                       "12h" = res[["12h"]] %>% 
left_join(targets, by = c("gene_id", "gene_name"), suffix = c(".stab", ".SNS")),
                       "24h" = res[["24h"]] %>% 
left_join(targets, by = c("gene_id", "gene_name"), suffix = c(".stab", ".SNS")))

## write the results tables to file
for (n in names(res_all_tables)) {
  write.table(res_all_tables[[n]], file = here("analysis", paste0(n, "_DE_stability.txt")),
              quote = FALSE, row.names = FALSE, sep = "\t")
}
```

Is the change in expression in the stability experiment the same as in the RNA-seq? We look at the logFC in the SNS RNA-seq and the stability RNA-seq.

```{r logFC-comparison-RNAseq}
for (n in names(res_tables)) {
  p <- ggplot(res_tables[[n]], aes(x = logFC.SNS,  y = logFC.stab)) + 
    geom_point(alpha = 0.8) + 
    theme_bw() + ggtitle(n)
  print(p)
}
```



# RNA stability over time {.tabset}

For all of the FUS targets, we plot the gene expression over all time points and genotypes.

```{r timeline-gene-expression, results = "asis", fig.width = 12}
## all genes that are DE for at least one time point (8h, 12h or 24h)
de_genes <- unique(unlist(lapply(res, function(x) x$gene_id)))

## get logcpms for all targets
g <- rownames(rowData(sg)[rowData(sg)$gene_id %in% de_genes,]) 
logcpms_part <- logcpms[g,]
## transform to long format
dat <- logcpms_part %>% as.data.frame %>% 
  rownames_to_column("gene_id") %>% 
  tidyr::gather(key = "sample_name", value = "logcpm", -gene_id) 
## merge with metadata to add time and genotype
dat <- dat %>% dplyr::left_join(metadata %>% as.data.frame, 
                                by = c("sample_name" = "names")) %>% 
  dplyr::rename(sample_id = sample_name)  %>% 
  dplyr::mutate(time = factor(time, levels = c("0h", "8h", "12h", "24h")))
dat$gene_name <- rowData(sg)[dat$gene_id, "gene_name"]
  
## table with the logFC per gene across all timepoints
dat_logfc <- res_all[[1]] %>% as.data.frame %>% 
  dplyr::select(gene_id, gene_name, logFC) %>% 
  dplyr::rename("0h" = logFC) %>% 
  dplyr::left_join(res_all[["8h"]] %>% as.data.frame %>% 
                     dplyr::select(gene_id, logFC), by = "gene_id") %>%
  dplyr::rename("8h" = logFC) %>%
  dplyr::left_join(res_all[["12h"]] %>% as.data.frame %>% 
                     dplyr::select(gene_id, logFC), by = "gene_id") %>%
   dplyr::rename("12h" = logFC) %>%
  dplyr::left_join(res_all[["24h"]] %>% as.data.frame %>%
                     dplyr::select(gene_id, logFC), by = "gene_id") %>%
   dplyr::rename("24h" = logFC) %>% 
  tidyr::gather(key = "time", value = "logFC", -gene_id, - gene_name) %>%
  dplyr::mutate(time = factor(time, levels = c("0h", "8h", "12h", "24h")))


## plot logcpms and logFC values of all targets among the DE
target_de <- targets$gene_name[targets$gene_id %in% de_genes]
for (g in target_de) {
  cat("## ", g, "\n")
  p1 <- ggplot(dat[dat$gene_name == g,], 
               aes(x = time, y = logcpm, color = genotype)) + 
    geom_point(size = 3, alpha = 0.8)  +
    theme_bw(base_size = 16)
  p2 <- ggplot(dat_logfc[dat_logfc$gene_name == g,], aes(x = time, y = logFC)) + 
    geom_point(size = 3, alpha = 0.8)  +
    theme_bw(base_size = 16)
  # lgd <- get_legend(p1)
  p <- plot_grid(p1, p2, nrow = 1, rel_widths = c(1, 1))
  print(p)
  ggsave(here("analysis", "figures", "per_gene", paste0(g, "_logcpms_logFC.png")), 
         p, width = 12, height = 6, dpi = 300, type = "cairo")
  cat("\n\n")
}
```

# Hierarchical clustering of DE genes

We cluster the DE genes based on logFC over all time points.

```{r clustering, warnings = FALSE}
## subset to DE genes
dat_logfc_de <- dat_logfc %>% dplyr::filter(gene_id %in% de_genes)
## we need a table with row = observations (genes) and columns = variables (time points)
logfc_wide <- dat_logfc_de %>% tidyr::spread(key = time, value = logFC)
gene_wide <- logfc_wide$gene_name
logfc_wide <- logfc_wide %>% dplyr::select(-gene_id, -gene_name) 
head(logfc_wide)
## distribution of logFCs per timepoint
apply(logfc_wide, 2, summary)

## compute distance metric
## euclidean distance
gene_dist <- dist(logfc_wide, method = "euclidean")
as.matrix(gene_dist)[1:6, 1:6]
## --> we don't want to use euclidean distance because we want to compare the logFC profiles and not so much the actual values

## correlation based distance metric
## the distance is 0 if two genes are perfectly correlated
## We want to identify clusters with similar profiles, independent of their magnitude (difference in y)
## pearson finds linear correlations, spearman is based on rank (monotonic relationship)
gene_dist <- get_dist(logfc_wide, method = "spearman")
round(as.matrix(gene_dist)[1:6, 1:6], 1)
gene_dist <- get_dist(logfc_wide, method = "pearson")
round(as.matrix(gene_dist)[1:6, 1:6], 1)


## scatterplot of logFC of two genes 
## change in same direction
plot(unlist(logfc_wide[gene_wide == "Syt1",]), unlist(logfc_wide[gene_wide == "Calm3",]))
cor(unlist(logfc_wide[gene_wide == "Syt1",]), unlist(logfc_wide[gene_wide == "Calm3",]), method = "pearson")
cor(unlist(logfc_wide[gene_wide == "Syt1",]), unlist(logfc_wide[gene_wide == "Calm3",]), method = "spearman")
cor(unlist(logfc_wide[gene_wide == "Syt1",]), unlist(logfc_wide[gene_wide == "Calm3",]), method = "kendall")
logfc_wide[gene_wide %in% c("Syt1", "Calm3"),] %>% dist(method = "euclidean")
logfc_wide[gene_wide %in% c("Syt1", "Calm3"),] %>% get_dist(method = "pearson")
logfc_wide[gene_wide %in% c("Syt1", "Calm3"),] %>% get_dist(method = "spearman")
logfc_wide[gene_wide %in% c("Syt1", "Calm3"),] %>% get_dist(method = "kendall")
## change in different direction
plot(unlist(logfc_wide[gene_wide == "Map2",]), unlist(logfc_wide[gene_wide == "Camk2a",]))
cor(unlist(logfc_wide[gene_wide == "Map2",]), unlist(logfc_wide[gene_wide == "Camk2a",]), method = "pearson")
cor(unlist(logfc_wide[gene_wide == "Map2",]), unlist(logfc_wide[gene_wide == "Camk2a",]), method = "spearman")
cor(unlist(logfc_wide[gene_wide == "Map2",]), unlist(logfc_wide[gene_wide == "Camk2a",]), method = "kendall")
logfc_wide[gene_wide %in% c("Map2", "Camk2a"),] %>% dist(method = "euclidean")
logfc_wide[gene_wide %in% c("Map2", "Camk2a"),] %>% get_dist(method = "pearson")
logfc_wide[gene_wide %in% c("Map2", "Camk2a"),] %>% get_dist(method = "spearman")
logfc_wide[gene_wide %in% c("Map2", "Camk2a"),] %>% get_dist(method = "kendall")

## We try spearman correlation
## Linkage function
hc <- hclust(d = gene_dist, method = "complete")

## dendrogram, its too big for visualization
# fviz_dend(hc, show_labels = FALSE, )

## compare tree height with original distance matrix
coph <- cophenetic(hc)
## correlation between cophenetic and original distance
cor(gene_dist, coph)

## single linkage
cor(gene_dist, cophenetic(hclust(gene_dist, method = "single")))

## does average linkage produce a better tree?
hc2 <- hclust(gene_dist, method = "average")
cor(gene_dist, cophenetic(hc2))
## yes it does

for (k in 2:8){
  hc_cut <- cutree(hc2, k = k)
  print(fviz_cluster(object = list(data = logfc_wide, cluster = hc_cut),
                     geom = "point", ellipse.type = "convex"))
  print(plot(silhouette(hc_cut, gene_dist), 
           main = paste("k = ", k), do.n.k=TRUE))
}

## cut the tree into groups
## What is the optimal number of clusters?
nbclust_all <- NbClust(data = logfc_wide, diss = gene_dist, distance = NULL, 
                       min.nc = 2, max.nc = 10, method = "average")
nbclust_all$Best.nc[1,]
nc_all <- sort(table(nbclust_all$Best.nc[1,]), decreasing=TRUE) %>% 
  names %>% as.numeric %>% .[1]
## number of genes per cluster
nbclust_all$Best.partition %>% table

## visualize clusters on PCA plot
hc_cut_all <- cutree(hc2, k = nc_all)
hc_cut_all %>% table

## plot the logFC per cluster
for (i in 1:nc_all){
  print(paste0("cluster ", i))
  ## select all genes in cluster
  p <-   dat_logfc_de %>% dplyr::filter(gene_name %in% 
                                          gene_wide[hc_cut_all$cluster == i]) %>%
    ggplot(aes(x = time, y = logFC, group = gene_name)) + 
    geom_line(color = "grey", alpha = 0.5) + 
    geom_point(size = 2, alpha = 0.6)  +
    theme_bw(base_size = 16) + 
    ggtitle(paste0("cluster ", i))
  print(p)
}
```


## All DE, different parameters {.tabset}

We test both pearson and spearman correlation and different linkage functions.
```{r clustering-DE-parameters, warnings = FALSE, message = FALSE, results = "asis", fig.width = 12}
test_clustering <- function(dat, dist, linkage, nc = 3, dat_logfc, gene_ids){
  gene_dist <- get_dist(dat, method = dist)
  tree <- hclust(gene_dist, method = linkage)
  cor(gene_dist, cophenetic(tree))
  hc_cut <- cutree(tree, k = nc)
  ## we want the cluster and silhouette plot in one row
  plot_list <- list()
  for (i in 1:nc){
    ## select all genes in cluster
    p <- dat_logfc %>% dplyr::filter(gene_name %in% 
                                            gene_ids[hc_cut == i]) %>%
      ggplot(aes(x = time, y = logFC, group = gene_name)) + 
      geom_line(color = "grey", alpha = 0.5) + 
      geom_point(size = 2, alpha = 0.6)  +
      theme_bw(base_size = 16) + 
      ggtitle(paste0(dist, " correlation; ", linkage, " linkage; cluster ", i))
    # print(p)
    plot_list[[i]] <- p
  }    
  print(plot_grid(plotlist = plot_list, ncol = 3))
  print(plot(silhouette(hc_cut, gene_dist), main = paste("k = ", i), do.n.k = TRUE))
  
}


dist_met <- c("pearson", "spearman")
link_met <- c("average", "complete", "centroid", "single", "ward.D2")
for(d in dist_met){
  for(n in 3:6){
    # par(mfrow=c(ceiling((n+1)/3), 3), by.row = TRUE)
    # pdf(here("analysis", "figures", "clustering", paste0("allDE_", d, "_", n, "cluster.pdf")))
    for(l in link_met){
      cat(paste0("### ", d, ": ",l, " linkage ", n, ", cluster \n"))
      test_clustering(logfc_wide, d, l, nc = n, dat_logfc, gene_wide)
      cat("\n\n")
    }
  }
}
```

# DE up
We might need to separate the genes into up and down DE, because the correlation based distance metric does not differentiate between positive and negative logFC. It is also hard to visualize such a big dendrogram.
```{r clustering-up, warnings = FALSE}
up <- de_genes <- unique(unlist(lapply(res, function(x) 
  x[x$FDR <= 0.05 & x$logFC > 0, "gene_id"])))
length(up)
## prepare data frame
dat_up <- dat_logfc %>% dplyr::filter(gene_id %in% up)

## we need a table with row = observations (genes) and columns = variables (time points)
dat_up <- dat_up %>% tidyr::spread(key = time, value = logFC)
gene_up <- dat_up$gene_name
dat_up <- dat_up%>% dplyr::select(-gene_id, -gene_name) 

## Clustering
gene_dist_up <- get_dist(dat_up, method = "pearson")
## linkage function
hc_up <- hclust(d = gene_dist_up, method = "complete")
## correlation between cophenetic and original distance
cor(gene_dist_up,  cophenetic(hc_up))
## single linkage
cor(gene_dist_up, cophenetic(hclust(gene_dist_up, method = "single")))

hc_up <- hclust(gene_dist_up, method = "average")
cor(gene_dist_up, cophenetic(hc_up))
## visualize clusters on PCA plot and silhouette width
for (k in 2:8){
  hc_cut_up <- cutree(hc_up, k = k)
  print(fviz_cluster(object = list(data = dat_up, cluster = hc_cut_up),
                     geom = "point", ellipse.type = "convex"))
  print(plot(silhouette(hc_cut_up, gene_dist_up), 
             main = paste("k = ", k), do.n.k=TRUE))
}

## what is the best number of clusters
nbclust_up <- NbClust(data = dat_up, diss = gene_dist_up, distance = NULL, 
                   min.nc = 2, max.nc = 8, method = "average")
## best number of clusters per index
nbclust_up$Best.nc[1,]
table(nbclust_up$Best.nc[1,])
nc_up <- sort(table(nbclust_up$Best.nc[1,]), decreasing=TRUE) %>% 
  names %>% as.numeric %>% .[1]
## number of genes per cluster
nbclust_up$Best.partition %>% table
bclust_up$Best.partition %>% table

## we use 5 clusters instead of 2, because it was the second best partition
nc_up <- 5
hc_cut_up <- cutree(hc_up, k = nc_up)
hc_cut_up %>% table

## plot the logFC per cluster
for (i in 1:nc_up){
  print(paste0("cluster ", i))
  ## select all genes in cluster
  p <-   dat_logfc_de %>% dplyr::filter(gene_name %in% 
                                          gene_up[hc_cut_up$cluster == i]) %>%
    ggplot(aes(x = time, y = logFC, group = gene_name)) + 
    geom_line(color = "grey", alpha = 0.5) + 
    geom_point(size = 2, alpha = 0.6)  +
    theme_bw(base_size = 16) + 
    ggtitle(paste0("cluster ", i))
  print(p)
}
```

## DE up, different paramters
```{r clustering-up-paramters, message = FALSE, warnings = FALSE}
dist_met <- c("pearson", "spearman")
link_met <- c("average", "complete", "centroid", "single", "ward.D2")
for(d in dist_met){
  for(n in 3:6){
    pdf(here("analysis", "figures", "clustering", paste0("DEup_", d, "_", n, "cluster.pdf")))
    for(l in link_met){
      test_clustering(dat_up, d, l, nc = n, dat_logfc_de, gene_up)
    }
    dev.off()
  }
}
```


# DE down
```{r clustering-down, warnings = FALSE}
down <- de_genes <- unique(unlist(lapply(res, function(x) 
  x[x$FDR <= 0.05 & x$logFC < 0, "gene_id"])))
length(down)
## prepare data frame
dat_down <- dat_logfc %>% dplyr::filter(gene_id %in% down)

## we need a table with row = observations (genes) and columns = variables (time points)
dat_down <- dat_down %>% tidyr::spread(key = time, value = logFC)
gene_down <- dat_down$gene_name
dat_down <- dat_down%>% dplyr::select(-gene_id, -gene_name) 

## Clustering
gene_dist_down <- get_dist(dat_down, method = "spearman")

## linkage function
hc_down <- hclust(d = gene_dist_down, method = "complete")
## correlation between cophenetic and original distance
cor(gene_dist_down,  cophenetic(hc_down))

## single linkage
cor(gene_dist_down, cophenetic(hclust(gene_dist_down, method = "single")))

hc_down <- hclust(gene_dist_down, method = "average")
cor(gene_dist_down, cophenetic(hc_down))

for (k in 2:8){
  hc_cut_down <- cutree(hc_down, k = k)
  print(fviz_cluster(object = list(data = dat_down, cluster = hc_cut_down),
                     geom = "point", ellipse.type = "convex"))
  print(plot(silhouette(hc_cut_down, gene_dist_down), 
             main = paste("k = ", k), do.n.k=TRUE))

}

## what is the best number of clusters
nbclust_down <- NbClust(data = dat_down, diss = gene_dist_down, distance = NULL, 
                   min.nc = 2, max.nc = 8, method = "average")
## best number of clusters per index
nbclust_down$Best.nc[1,]
table(nbclust_down$Best.nc[1,])
nc_down <- sort(table(nbclust_down$Best.nc[1,]), decreasing=TRUE) %>% 
  names %>% as.numeric %>% .[1]
## number of genes per cluster
nbclust_down$Best.partition %>% table

hc_cut_down <- cutree(hc_down, k = nc_down)
hc_cut_down %>% table

## plot the logFC per cluster
for (i in 1:nc_down){
  print(paste0("cluster ", i))
  ## select all genes in cluster
  p <-   dat_logfc_de %>% dplyr::filter(gene_name %in% 
                                          gene_down[hc_cut_down$cluster == i]) %>%
    ggplot(aes(x = time, y = logFC, group = gene_name)) + 
    geom_line(color = "grey", alpha = 0.5) + 
    geom_point(size = 2, alpha = 0.6)  +
    theme_bw(base_size = 16) + 
    ggtitle(paste0("cluster ", i))
  print(p)
}
```

## DE down, different parameters
```{r clustering-down-paramters, message = FALSE, warnings = FALSE}
dist_met <- c("pearson", "spearman")
link_met <- c("average", "complete", "centroid", "single", "ward.D2")
for(d in dist_met){
  for(n in 3:6){
    pdf(here("analysis", "figures", "clustering", paste0("DEdown_", d, "_", n, "cluster.pdf")))
    for(l in link_met){
      test_clustering(dat_down, d, l, nc = n, dat_logfc_de, gene_down)
    }
    dev.off()
  }
}
```


# All FUS targets
```{r clustering-all-targets, warnings = FALSE}
targets_id <- targets$gene_id
length(targets_id)
## prepare data frame
dat_fus <- dat_logfc %>% dplyr::filter(gene_id %in% targets_id) %>% 
  dplyr::filter(!is.na(logFC))

## we need a table with row = observations (genes) and columns = variables (time points)
dat_fus <- dat_fus %>% tidyr::spread(key = time, value = logFC)
gene_fus <- dat_fus$gene_name
dat_fus <- dat_fus%>% dplyr::select(-gene_id, -gene_name) 

## Clustering
gene_dist_fus <- get_dist(dat_fus, method = "spearman")

## linkage function
hc_fus <- hclust(d = gene_dist_fus, method = "complete")
## correlation between cophenetic and original distance
cor(gene_dist_fus,  cophenetic(hc_fus))
## single linkage
cor(gene_dist_fus, cophenetic(hclust(gene_dist_fus, method = "single")))
## average linkage
hc_fus <- hclust(gene_dist_fus, method = "average")
cor(gene_dist_fus, cophenetic(hc_fus))

for (k in 2:8){
  hc_cut_fus <- cutree(hc_fus, k = k)
  print(fviz_cluster(object = list(data = dat_fus, cluster = hc_cut_fus),
                     geom = "point", ellipse.type = "convex"))
  print(plot(silhouette(hc_cut_fus, gene_dist_fus), 
           main = paste("k = ", k), do.n.k=TRUE))
}

## what is the best number of clusters
nbclust_fus <- NbClust(data = dat_fus, diss = gene_dist_fus, distance = NULL, 
                   min.nc = 2, max.nc = 8, method = "average")
## best number of clusters per index
nbclust_fus$Best.nc[1,]
table(nbclust_fus$Best.nc[1,])
nc_fus <- sort(table(nbclust_fus$Best.nc[1,]), decreasing=TRUE) %>% 
  names %>% as.numeric %>% .[1]
## number of genes per cluster
nbclust_fus$Best.partition %>% table

hc_cut_fus <- cutree(hc_fus, k = nc_fus)
hc_cut_fus %>% table

## plot the logFC per cluster
for (i in 1:nc_fus){
  print(paste0("cluster ", i))
  ## select all genes in cluster
  p <-   dat_logfc %>% dplyr::filter(gene_name %in% 
                                          gene_fus[hc_cut_fus$cluster == i]) %>%
    ggplot(aes(x = time, y = logFC, group = gene_name)) + 
    geom_line(color = "grey", alpha = 0.5) + 
    geom_point(size = 2, alpha = 0.6)  +
    theme_bw(base_size = 16) + 
    ggtitle(paste0("cluster ", i))
  print(p)
}
```


## Fus targets, different parameters

```{r clustering-fus-targets-paramters, warnings = FALSE, message = FALSE}
dist_met <- c("pearson", "spearman")
link_met <- c("average", "complete", "centroid", "single", "ward.D2")
for(d in dist_met){
  for(n in 3:6){
    pdf(here("analysis", "figures", "clustering", paste0("FUStargets_", d, "_", n, "cluster.pdf")))
    for(l in link_met){
      
      test_clustering(dat_fus, d, l, nc = n, dat_logfc, gene_fus)
    }
    dev.off()
  }
}
```

## FUS targets not DE

We separately cluster the FUS targets that are not DE at any time point.

```{r clustering-FUS-notDE}
## all FUS targets not in DE stability
targets_notDE_id <- targets_id[!targets_id %in% c(up, down)]
length(targets_notDE_id)
## prepare data frame
dat_fus <- dat_logfc %>% dplyr::filter(gene_id %in% targets_notDE_id) %>% 
  dplyr::filter(!is.na(logFC))

## we need a table with row = observations (genes) and columns = variables (time points)
dat_fus <- dat_fus %>% tidyr::spread(key = time, value = logFC)
gene_fus <- dat_fus$gene_name
dat_fus <- dat_fus%>% dplyr::select(-gene_id, -gene_name) 

## Clustering
gene_dist_fus <- get_dist(dat_fus, method = "pearson")

## linkage function
hc_fus <- hclust(d = gene_dist_fus, method = "complete")
## correlation between cophenetic and original distance
cor(gene_dist_fus,  cophenetic(hc_fus))

## single linkage
cor(gene_dist_fus, cophenetic(hclust(gene_dist_fus, method = "single")))

hc_fus <- hclust(gene_dist_fus, method = "average")
cor(gene_dist_fus, cophenetic(hc_fus))

for (k in 2:8){
  hc_cut_fus <- cutree(hc_fus, k = k)
  print(fviz_cluster(object = list(data = dat_fus, cluster = hc_cut_fus),
                     geom = "point", ellipse.type = "convex"))
  print(plot(silhouette(hc_cut_fus, gene_dist_fus), 
             main = paste("k = ", k), do.n.k=TRUE))
}

## what is the best number of clusters
nbclust_fus <- NbClust(data = dat_fus, diss = gene_dist_fus, distance = NULL, 
                   min.nc = 2, max.nc = 8, method = "average")
## best number of clusters per index
nbclust_fus$Best.nc[1,]
table(nbclust_fus$Best.nc[1,])
nc_fus <- sort(table(nbclust_fus$Best.nc[1,]), decreasing=TRUE) %>% 
  names %>% as.numeric %>% .[1]
## number of genes per cluster
nbclust_fus$Best.partition %>% table

hc_cut_fus <- cutree(hc_fus, k = nc_fus)
hc_cut_fus %>% table

## plot the logFC per cluster
for (i in 1:nc_fus){
  print(paste0("cluster ", i))
  ## select all genes in cluster
  p <-   dat_logfc %>% dplyr::filter(gene_name %in% 
                                          gene_fus[hc_cut_fus == i]) %>%
    ggplot(aes(x = time, y = logFC, group = gene_name)) + 
    geom_line(color = "grey", alpha = 0.5) + 
    geom_point(size = 2, alpha = 0.6)  +
    theme_bw(base_size = 16) + 
    ggtitle(paste0("cluster ", i))
  print(p)
}
```


## Fus targets not DE, different parameters

```{r clustering-fus-notDE-paramters, warnings = FALSE, message = FALSE}
dist_met <- c("pearson", "spearman")
link_met <- c("average", "complete", "centroid", "single", "ward.D2")
for(d in dist_met){
  for(n in 3:6){
    pdf(here("analysis", "figures", "clustering", paste0("FUStargets_notDE_", d, "_", n, "cluster.pdf")))
    for(l in link_met){
      
      test_clustering(dat_fus, d, l, nc = n, dat_logfc, gene_fus)
    }
    dev.off()
  }
}
```



# location of FUS binding

We want to investigate whether genes that are DE up or down or unchanged in stability in KI vs. WT have different FUS binding patterns. We split the FUS targets into these 3 classes and plot the distribution of peak locations on pre-mRNA.

We prepare the annotation and read the list of synaptoneurosome FUS peaks after MA-plot filtering.
```{r prepare-peaks-anno, warnings = FALSE}
source(here("scripts", "data_import.R"))

## read GTF and split by annotation
anno <- prepare_anno(gtf)

## SNS FUS peaks 
peaks <- readRDS(fus_peaks_rds)
peaks <- peaks[peaks$gene_name %in% targets$gene_name]
```

We split the peaks into the three groups and plot the distribution of their annotated location.
```{r peaks-annotation-barplot}
## split peaks in three sets
fus_gene_list <- list(up = up[up %in% targets$gene_id], 
                       down = down[down %in% targets$gene_id], 
      unchanged = targets$gene_id[!(targets$gene_id %in% unique(c(up, down)))])
lengths(fus_gene_list)

## separate the peaks into sets
fus_peak_list <- list(up = peaks[peaks$gene_id %in% fus_gene_list[["up"]]],
                      down = peaks[peaks$gene_id %in% fus_gene_list[["down"]]],
            unchanged = peaks[peaks$gene_id %in% fus_gene_list[["unchanged"]]])
lengths(fus_peak_list)

## split the peaks by gene set
peak_gene_region_barplot(fus_peak_list, anno)
```

